Reproducible Research - Peer Assessment 1
========================================================

Below you'll find the relevant code and explanations for the first assignment of the Reproducible Research data science class.

### Loading and preprocessing the data

The following loads necessary packages for answering the assignment questions
```{r libraries, echo = TRUE}
library(knitr)
library(reshape2)
library(lattice)
```

Loads the assignment dataset 
```{r loaddata, echo = TRUE}
data <- read.csv("C:/Users/Alan/Desktop/RWD/activity.csv")
```

### What is the mean number of steps taken each day?

Create and transform new dataframe for plotting
```{r, echo = TRUE}
data2 <- melt(data, id = c("date", "interval"))
data2 <- dcast(data2, date ~ variable, fun.aggregate = sum)
```

Create histogram on transformed data
```{r fig.width = 7, fig.height = 6, echo = TRUE}
hist(data2$steps, breaks = 20)
```

Calculate mean and median values of total steps by day
```{r, echo = TRUE}
mean(data2$steps, na.rm = TRUE)
median(data2$steps, na.rm = TRUE)
```

### What is the average daily activity pattern?

Create and transform new dataframe for plotting
```{r, echo = TRUE}
data3 <- melt(data, id = c("date", "interval"))
data3 <- dcast(data3, interval ~ variable, fun.aggregate = mean, na.rm = TRUE)
```

Create time series plot on transformed data
```{r fig.width = 7, fig.height = 6, echo = TRUE}
with(data3, plot(interval, steps, type = "l",
                 ylab = "average steps across all days", xlab = "interval"))
```

Order data to determine interval with max average step value (interval 835)
```{r, echo = TRUE}
data3 <- data3[order(data3$steps, decreasing = TRUE), ]
data3[1,]
```

### Imputing missing values

Calculate total number of rows with NAs
```{r, echo = TRUE}
nrow(data[is.na(data$steps)==TRUE,])
```

Create new dataframe with NAs replaced with average steps per interval across days
```{r, echo = TRUE}
data4 <- data
for (i in c(1:nrow(data4))){
    if(is.na(data4$steps[i])){
        data4$steps[i] <- data3$steps[data3$interval==data4$interval[i]]
    }
}
```

Create and transform new dataframe based on newly imputed dataframe for plotting
```{r, echo = TRUE}
data5 <- melt(data4, id = c("date", "interval"))
data5 <- dcast(data5, date ~ variable, fun.aggregate = sum)
```

Create histogram on newly imputed dataframe
```{r fig.width = 7, fig.height = 6, echo = TRUE}
hist(data5$steps, breaks = 20)
```

Calculate mean and median values of total steps by day.  These new values differ from results calculated from the non-imputed dataframe.  Imputing strategy has resulted in raising the median to match the mean.
```{r, echo = TRUE}
mean(data5$steps, na.rm = TRUE)
median(data5$steps, na.rm = TRUE)
```

### Are there differences in activity patterns bewteen weekdays and weekends?

Create new dataframe with weekday/weekend factor variable
```{r, echo = TRUE}
data6 <- data4
for (i in c(1:nrow(data6))){
    if(weekdays(as.Date(data6$date[i])) %in% c("Saturday", "Sunday")){
        data6$w[i] <- "weekend"
    } else {
        data6$w[i] <- "weekday"
    }
}
```

Create and transform new dataframe with weekday/weekend factor
```{r, echo = TRUE}
data7 <- melt(data6, id = c("date", "interval", "w"))
data7 <- dcast(data7, interval + w ~ variable, fun.aggregate = mean, na.rm = TRUE)
```

Create panel plot by weekday type 
```{r fig.width = 7, fig.height = 6, echo = TRUE}
xyplot(steps ~ interval | w, data7, type = "l", layout = c(1,2))
```